Recovery of mountain plant communities in response to reductions in Nitrogen emissions is hidden by other drivers of global change

### 2018-06-12 Manuscript.Rmd
Update Manuscript.Rmd
Ellenberg ersetzt
Introduction eingefügt
Diskussion

# Introduction {-}
Nitrogen (N) deposition is a major threat to biodiversity. The recent introduction of cleaner technologies has let to reductions in the emissions of nitrogen oxides, with affiliated decrease in N-deposition in many parts of Europe. However, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of extant plant communities.

One useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both immigration and extinction, often in a closed range of values (Hillebrand_et_al-2018.pdf).

Here we inferred mountain hay meadows (EUNIS E2.3) in Switzerland. Explain why mountain hay meadows are important. Also explain other threats to mountain hay meadows (climate change, land-use change).

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hey communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the annual mean temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual mean precipitation per site from the WordClim database [@Fick2017].Further, we estimated atmospheric N deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of N deposition see @Rihm2001 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed. (folgenden Satz streichen?) We expect that well-accessible sites become more intensively managed, whereas management intensity will rather decrease at remote sites [@Strebel2015].

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Ellenberg value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average Ellenberg value of species that (a) disappeared from site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear model with the difference between disappeard/colonized species and random species as dependend variable and the sitemeasure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

@@

# Discussion {-}
- *General points*: Although N deposition considerably declined between 2005 and 2015, we could not detect major shifts in plant community structure during the same time period.

@@

- *Methodological point*: The rather large spatial turnover might be partly explained by species that remained undetected in one of the surveys. However, our results suggest that turnover is caused at least partly by species with specific Ellenberg values. These deviation from what we would expect under random species turnover is unlikely to be explained by species that remained undetected.

>>>>

# Introduction {-}
Nitrogen deposition is supposed to be one of the major threats to biodiversity (Sala 2000, Bobbink 2010, Menier et al. 2016). There is strong evidence for the impairment of biodiversity by nitrogen deposition for species-rich acid and calcareous grasslands in Europe (Stevens et al. 2006, Maskell et al. 2010, Duprè et al. 2010). Fertilizing experiments show that plant biodiversity in mountain grasslands is negatively affected by nitrogen addition, too (Humbert et al. 2016). 
Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe (Tørseth et al., 2012, Maas and Grennfelt, 2016) [Tørseth et al.,: Atmos. Chem. Phys., 12, 5447–5481, 2012].

However, it is an open question whether and how fast the reduction in N deposition rates will lead to the recovery of extant plant communities.  Quantifying the potential recovery of biodiversity in response to reducing air pollution requires an alternative to the space-for-time approach, ideally monitoring long-term community dynamics on permanent plots (Duprã, C. et al. Changes in species richness and composition in European acidic grasslands over the past 70 years: the contribution of cumulative atmospheric nitrogen deposition. Glob. Change Biol. 16, 344–357, (2010).

One useful approach to understanding biodiversity change is through estimates of biodiversity turnover reflecting both immigration and extinction, often in a closed range of values (Hillebrand et al-2018.pdf).

Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows (Roth et al. 2013). 
In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows of the Swiss mountain region (Schlup et al. 2013). In the alpine region, meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006).
The meta-analysis of Humbert et al. (2016) showed management and abiotic factors to interact with nitrogen addition to drive plant community changes: The overall effects of N addition on plant species richness and diversity (Shannon index) were more pronounced in managed experimental sites with warmer summers. Management regimes of accessible mountainous areas are intensified, while poorly accessible mountainous areas are being abandoned (Tasser and Tappeiner, 2002; Graf et al., 2014, Strebel and Bühler, 2015). This change in traditional management regimes is considered as a major driver that endangers grassland biodiversity in mountainous regions of Europe (Niedrist et al., 2009; Homburger and Hofer, 2012).

On European mountain summits, plants shifted upwards in the last decades (Rumpf et al. 2018). The authors conclude that «These results suggest that recent climate warming interacted with airborne nitrogen deposition in driving the observed dynamics. As changes increased more for species from lower elevations and eutrophic plant species showed the strongest increase in abundance, these findings might be relevant for mountain hay meadows, too.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the annual mean temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual mean precipitation per site from the WordClim database [@Fick2017].Further, we estimated atmospheric N deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of N deposition see @Rihm2001 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed. (folgenden Satz streichen?) We expect that well-accessible sites become more intensively managed, whereas management intensity will rather decrease at remote sites [@Strebel2015].

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Ellenberg value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear model with the difference between disappeard/colonized species and random species as dependend variable and the sitemeasure (gradient) as predictor variable.") %>% 
  kable_styling() %>% 
  group_rows("(a) Plants that disappeard from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

@@

# Discussion {-}
- *General points*: Although N deposition considerably declined between 2005 and 2015, we could not detect major shifts in plant community structure during the same time period. 
Although NOx emissions in Switzerland decreased by 46% between 1990 and 2010 and NH3 emissions by 14% (Maas and Grennfelt, 2016), nitrogen deposition on our plots decreased by only 2.7 kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015 (Meteotest, 2018). This is only about one tenth of the decrease in England in the same period, where Nitrogen deposition decreased by 24kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 (Storkey et al. 2015). In 2015, average N deposition was 14.8 kg ha$^{-1}$ yr$^{-1}$, about 85% of 2000 deposition, and due to the rather short period of time and the small decrease in nitrogen deposition, it is not surprising that species richness has not reacted significantly.

@@

- *Methodological point*: The rather large spatial turnover might be partly explained by species that remained undetected in one of the surveys. From the randomly selected quality control of the BDM 17 mountain hay meadows were investigated from 2003 to 2017. Species richness was surveyed in the same year by two independent botanists. The species match is (mean ± SD) 85 ± 5.2 per cent, suggesting that at least a portion of the turn over is due to agent effects. However, our results suggest that turnover is caused at least partly by species with specific indicator values. This deviation from what we would expect under random species turnover is unlikely to be explained by species that remained undetected.
